A Systematic Literature Review on Extensions to Role-Based Access Control

### 2013-05-08 paper/sections/abstract.tex
Updates to address grammatical issues.

\noindent
{\bf Context}: Since the National Institute of Standards and Technology (NIST) proposed Role-Based Access Control (RBAC) standard using RBAC reference models (RBAC$_{m}$ in short) in the late 1990's, computing in new domains has led to proposals for extensions to the RBAC$_{m}$; for example adding context around permission granting and role activation.
\\
\\
\noindent
{\bf Aim}: The goal of our work is to aid practitioners and researchers in choosing an extension to the RBAC$_{m}$, and in understanding
how extensions to RBAC$_{m}$ are evaluated by providing an assessment of the state of extensions to the RBAC$_{m}$. We accomplished this through: 
establishment of a set of extension categories, an examination of the state of the art in evaluations of extension models, and a breakdown of the 
motivations that have led to extension of the RBAC$_{m}$.
\\
\\
\noindent
{\bf Method}: We performed a systematic literature review that yielded 1,716 papers, of which 28 were deemed primary sources for inclusion as extensions to the RBAC$_{m}$. 
\\
\\
\noindent
{\bf Results}: Our results show that extensions to the RBAC$_{m}$ can be classified under eight categories: Constraint, Context, 
Organization, Privacy, Task, Spatio-Temporal, Spatial, and Temporal. We identified only
eight of the 28 papers provided an implementation of their model in the form of an enterprise application or prototype. We found that the primary domains 
that inspired extensions were the medical domain with nine of the 28 models, enterprise workflows with five of 28, and mobile computing with five of 28.
\\
\\
\noindent
{\bf Conclusions}: When examined as an entire group, the eight categories can be aligned under the single category of context. 
Our literature review shows that the state of evaluation of RBAC model evaluation is limited with one
paper doing an evaluation comparing itself to the RBAC$_{m}$, eight presenting example 
scenarios of their model in action, and 12 of the 28 models lacking an evaluation that we could discern. The landscape of extensions
to the RBAC$_{m}$ serves as a basis from which a new, and vetted, extended RBAC reference models should be derived.

>>>>

\noindent
{\bf Context}: Since the National Institute of Standards and Technology (NIST) proposed Role-Based Access Control (RBAC) standard using RBAC reference models (RBAC$_{m}$ in short) in the late 1990's, computing in new domains has led to proposals for extensions to the RBAC$_{m}$; for example adding context around permission granting and role activation.
\\
\\
\noindent
{\bf Aim}: The goal of our work is to aid practitioners and researchers in choosing an extension to the RBAC$_{m}$, and in understanding
how extensions to RBAC$_{m}$ are evaluated by providing an assessment of the state of extensions to the RBAC$_{m}$. We accomplished this through
establishment of a set of extension categories, an examination of the state of the art in evaluations of extension models, and a breakdown of the 
motivations that have led to extension of the RBAC$_{m}$.
\\
\\
\noindent
{\bf Method}: We performed a systematic literature review that yielded 1,716 papers, of which 28 were deemed primary sources for inclusion as extensions to the RBAC$_{m}$. 
\\
\\
\noindent
{\bf Results}: Our results show that extensions to the RBAC$_{m}$ can be classified under eight categories: Constraint, Context, 
Organization, Privacy, Task, Spatio-Temporal, Spatial, and Temporal. We identified only
8 of the 28 papers provided an implementation of their model in the form of an enterprise application or prototype. We found that the primary domains 
that inspired extensions were the medical domain with 9 of the 28 models, enterprise workflows with 5 of 28, and mobile computing with 5 of 28.
\\
\\
\noindent
{\bf Conclusions}: When examined as an entire group, the eight categories can be aligned under the single category of context. 
Our literature review shows that the state of evaluation of RBAC model evaluation is limited with 1
paper doing an evaluation comparing itself to the RBAC$_{m}$, 8 presenting example 
scenarios of their model in action, and 12 of the 28 models lacking an evaluation that we could discern. The landscape of extensions
to the RBAC$_{m}$ serves as a basis from which a new, and vetted, extended RBAC reference models should be derived.

### 2013-05-08 paper/sections/analysis.tex
Updates to address grammatical issues.

Table \ref{tab:implementations} shows the breakdown of implementations found within the primary sources.
Of the 28 papers surveyed, there was a lack of implementation with 20 of the paper providing no
mention of an implementation or prototype.  Of the remaining eight papers that did mention an implementation, half 
were simply prototypes developed by the authors while the other half were claimed to be implemented within a real
system.

@@

Based on the diverse evaluation criteria, 12 models presented no evidence of an evaluation. Eight models presented example scenarios
and how application of their model would apply and resolve the situation.  Six of the models provided some form of performance
or complexity analysis of their model.  This included graphs of the model's time to determine authorization as the number of entities
grew, and the size of the role space for the extension model compared to standard RBAC. Four models provided mathematical descriptions
and analysis as a way to provide evaluation in the form of completeness. 
The most widely used evaluation method was providing sample scenarios with accompanying workflows of how the extension model
would tackle those scenarios. Much is left to the reader to assume of these types of evaluations, as the authors do not explicitly state
or show how the RBAC$_{m}$ is deficient in tackling said scenarios.

@@

The predominant domain for which extension models have been generated for is that of the medical domain with nine of 28 mentioning scenarios or requirements of that industry.
Mobile computing and enterprise workflows were each represented by five papers claiming to be influenced by the requirements for access control within these domains. The final set
of domains was pervasive computing environments and large-scale organizations with three each and web applications with one.  There were four papers without any direct mention of a domain
since Aich et al. \cite{aich09:role} fall under both the medical domain and mobile computing.

@@

The RBAC standard has been re-enforced by the economic impact that standardization has had on enterprises needing to apply access control.  The
inclusion of extension models targeted at the enterprise workflow domain is indicative of the expansion of requirements for enterprises. Developers
and researchers should take care when looking at extension models designed to addresses the newer requirements of enterprise workflows in order to
achieve the same economic implementation and maintainability benefits the RBAC$_{m}$ presents.

>>>>

Table \ref{tab:implementations} shows the breakdown of implementations found within the primary sources.
Of the 28 papers surveyed, there was a lack of implementation with 20 of the paper providing no
mention of an implementation or prototype.  Of the remaining eight papers that did mention an implementation, half 
were simply prototypes developed by the authors whereas the other half were claimed to be implemented within a real
system.

@@

Based on the diverse evaluation criteria, 12 models presented no evidence of an evaluation. Eight models presented example scenarios
and how application of their model would apply and resolve the situation.  Six of the models provided some form of performance
or complexity analysis of their model.  The performance and complexity analysis included graphs of the model's time to determine calculate authorization 
as the number of entities grew, and the size of the role space for the extension model compared to the RBAC Reference Model. 
Four models provided mathematical descriptions and analysis as a way to provide evaluation in the form of completeness. 
The most widely used evaluation approach was providing sample scenarios with accompanying workflows of how the extension model
would tackle those scenarios. Much is left to the reader to assume of these types of evaluations, as the authors do not explicitly state
or show how the RBAC$_{m}$ is deficient in tackling said scenarios.

@@

The predominant domain for which extension models have been generated for is that of the medical domain with 9 of 28 mentioning scenarios or requirements of that industry.
Mobile computing and enterprise workflows were each represented by five papers claiming to be influenced by the requirements for access control within these domains. The final set
of domains was pervasive computing environments and large-scale organizations with three each and web applications with one.  There were four papers without any direct mention of a domain
since Aich et al. \cite{aich09:role} fall under both the medical domain and mobile computing.

@@

The RBAC standard has been re-enforced by the economic impact that standardization has had on enterprises needing to apply access control.  The
inclusion of extension models targeted at the enterprise workflow domain is indicative of the expansion of requirements for enterprises. Developers
and researchers should take care when looking at extension models designed to address the newer requirements of enterprise workflows in order to
achieve the same economic implementation and maintainability benefits the RBAC$_{m}$ presents.

### 2013-05-08 paper/sections/categorization.tex
pdf upload and minior update
Updates to address grammatical issues.

During the paper reading phase, we identify categorical themes, which are
central topic or context of RBAC extensions. 
For example, the paper ``Privacy-aware role-based access control'' \cite{ni2010privacy} brings in the notion of privacy explicitly within the title of the paper and the name of their model. 
Some papers presented this direct pronouncement of theme of RBAC extensions, while others were less obvious. 
Thus, we developed a set of guidelines to aide in determining a set of eight categories based upon observations during the paper reading phase. 
In developing these guidelines, we defined each category by a single noun-phrase descriptor.
The guidelines were:

@@

  \item \textbf{Privacy (Priv)}: The extension model provides entities and mechanisms to describe privacy policies, which are legal statements or documents about disclosure or management of personally identifiable information such as name, address, data of birth, etc.

  \item \textbf{Task}: The extension model provides task elements which are associated with permissions and roles. A task is a fundamental unit of a business activity. Different from core RBAC, in task-role-based access control model, roles are not directly associated with permissions. Roles are associated with tasks that are associated with permissions. For example, the employee role is associated with a task, for example, to write a report. Then, this task is associated with a permission.

@@

  \item \textbf{Spatial}: The extension model provides spatial constraints, location-based constraints in specifying access
	control policies. For example, in organizations, locations are enforced while a
	specific role is permitted to conduct an action. Consider that a part-time employee works only in a specific location.
	In such cases, the part-time employee role should access required resources only when the user is in that location. 
	Spatial constraints can integrate with roles, user-role assignments, or role-permission assignments.

  \item \textbf{Temporal (Temp)}:  The extension model provides temporal constraints, time-based constraints in specifying access
	control policies. For example, in organizations, periodic temporal durations are enforced while a
	specific role is permitted to conduct an action. Consider that a part-time employee works only from 9:00 a.m. to 3:00 p.m.
	In such cases, the part-time employee role should be allowed to access required resources during the interval. 
	Temporal constraints can integrate with roles, user-role assignments, or role-permission assignments.

@@

When looking across all categories, the authors noted that each category added some new features on top of the standard RBAC model that were domain specific.
These domain specific features were under the surface these features adding contextual relationships between the core user, permission and role entities.  Thus, the authors
concluded that all categories stemmed from the context category, of which some primary sources were already deemed direct members.

For example, in the case of the Privacy category, the models added entities such as purpose binding to represent within the model data collected for one purpose should not used for another purpose without user consent {\cite{ni2010privacy}. 
While the new entity provided by the Privacy based models is inspired by domains such as healthcare where privacy is of legal concern, the underlying mechanism that drives purpose binding is providing context around making an access control decision. 
The system must take into account not just a static set of permissions a user has through their roles, but also the context of the data being accessed as that data relates to privacy policy. 
In the spatio-temporal models, a users location and the time of day are two factors that can be taken into account when activating a role or verifying a permission.  
The concepts of location and time are properties of the user and place specific contexts around the role and permission entities.

>>>>

During the paper reading phase, we identify categorical themes, which are
central topic or context of RBAC extensions. 
For example, the paper ``Privacy-aware role-based access control'' \cite{ni2010privacy} brings in the notion of privacy explicitly within the title of the paper and the name of their model. 
Some papers presented this direct pronouncement of theme of RBAC extensions, whereas others were less obvious. 
Thus, we developed a set of guidelines to aide in determining a set of eight categories based upon observations during the paper reading phase. 
In developing these guidelines, we defined each category by a single noun-phrase descriptor.
The guidelines were

@@

  \item \textbf{Privacy (Priv)}: The extension model provides entities and mechanisms to describe privacy policies, which are legal statements or documents about disclosure or management of personally identifiable information such as name, address, or data of birth.

  \item \textbf{Task}: The extension model provides task entities which are associated with permissions and roles. A task is a fundamental unit of a business activity. Different from core RBAC, in task-role-based access control model, roles are not directly associated with permissions. Roles are associated with tasks that are associated with permissions. For example, the employee role is associated with a task, for example, to write a report. Then, this task is associated with a permission.

@@

  \item \textbf{Spatial}: The extension model provides spatial constraints, location-based constraints in specifying access
	control policies. For example, in organizations, locations are enforced whereas a
	specific role is permitted to conduct an action. Consider that a part-time employee works only in a specific location.
	In such cases, the part-time employee role should access required resources only when the user is in that location. 
	Spatial constraints can integrate with roles, user-role assignments, or role-permission assignments.

  \item \textbf{Temporal (Temp)}:  The extension model provides temporal constraints, time-based constraints in specifying access
	control policies. For example, in organizations, periodic temporal durations are enforced whereas a
	specific role is permitted to conduct an action. Consider that a part-time employee works only from 9:00 a.m. to 3:00 p.m.
	In such cases, the part-time employee role should be allowed to access required resources during the interval. 
	Temporal constraints can integrate with roles, user-role assignments, or role-permission assignments.

@@

When looking across all categories, we noted that each category added some new features on top of the standard RBAC model that were domain specific.
These domain specific features were under the surface these features adding contextual relationships between the core user, permission and role entities.  Thus, we
concluded that all categories stemmed from the context category, of which some primary sources were already deemed direct members.

For example, in the case of the Privacy category, the models added entities such as purpose binding to represent within the model data collected for one purpose should not be used for another purpose without user consent {\cite{ni2010privacy}. 
While the new entity provided by the Privacy based models is inspired by domains such as healthcare where privacy is of legal concern, the underlying mechanism that drives purpose binding is providing context around making an access control decision. 
The system must take into account not just a static set of permissions a user has through their roles, but also the context of the data being accessed as that data relates to privacy policy. 
In the spatio-temporal models, a users location and the time of day are two factors that can be taken into account when activating a role or verifying a permission.  
The concepts of location and time are properties of the user and place specific contexts around the role and permission entities.

### 2013-05-08 paper/sections/core_rbac.tex
pdf upload and minior update
Updates to address grammatical issues.

Figure~\ref{fig:overview} presents an overview of core RBAC reference model diagram where elements and their relations are described.
Let "USERS", "ROLES", "OBS", "OPS", "PRM", and "SESSIONS" denote users, roles, objects, operations, permissions, and sessions, respectively.
Permissions are associated with possible users' pre-defined operation on an object (e.g., execute a file).
Note that, at user or role activation, a session associated with user or role is established.

@@

The hierarchical RBAC reference model adds role hierarchies ($RH$) feature to the core RBAC reference model.
$RH$ incorporates a structure of roles in an organization using inheritance relationships among attributes such as roles.
The role structure in an organization may use
a role $r_1$, which inherits all permissions of another role $r_2$.
For example, a manager role may inherit all permissions of a employee role.
Role hierarchies help simplify access control policy creation and maintenance by reducing the number of
individual role assignments for a user. Formally, the role inheritance relation is shown as \textit{RH} $\subseteq$ \textit{ROLES} $\times$ \textit{ROLES} describing the many-to-many mapping role inheritance relation. 
General role hierarchies can be extended to use the concept of multiple inheritances where
$r_1$ inherits all permissions from more than one roles.

@@

The constrained RBAC reference model incorporates separation of duty relations to  the core or the hierarchical RBAC reference. Separation of 
duty relations enforces conflict of interest among roles. This model defines two types of separation of duty relations; static and dynamic.

>>>>

Figure~\ref{fig:overview} presents an overview of core RBAC reference model diagram where entities and their relations are described.
Let "USERS", "ROLES", "OBS", "OPS", "PRM", and "SESSIONS" denote users, roles, objects, operations, permissions, and sessions, respectively.
Permissions are associated with possible users' pre-defined operation on an object (e.g., execute a file).
Note that, at user or role activation, a session associated with user or role is established.

@@

The hierarchical RBAC reference model adds role hierarchies ($RH$) feature to the core RBAC reference model.
$RH$ incorporates a structure of roles in an organization using inheritance relationships among attributes such as roles.
The role structure in an organization may use
a role $r_1$, which inherits all permissions of another role $r_2$.
For example, a manager role may inherit all permissions of an employee role.
Role hierarchies help simplify access control policy creation and maintenance by reducing the number of
individual role assignments for a user. Formally, the role inheritance relation is shown as \textit{RH} $\subseteq$ \textit{ROLES} $\times$ \textit{ROLES} describing the many-to-many mapping role inheritance relation. 
General role hierarchies can be extended to use the concept of multiple inheritances where
$r_1$ inherits all permissions from more than one roles.

@@

The constrained RBAC reference model incorporates separation of duty relations to  the core or the hierarchical RBAC reference. Separation of 
duty relations enforces conflict of interest among roles. This model defines two types of separation of duty relations: static and dynamic.

### 2013-05-08 paper/sections/discussion.tex
Updates to address grammatical issues.

This standard does not provide a specific guideline for implementation of various extended RBAC models.
Moreover, when software developers implemented extended RBAC models, they require clear security requirements of extended role-based access control, which identifies security objectives, intended environments, and assumed threats. Moreover, software developers understand how this extended RBAC overlaps with other access control models when more than two access control models are integrated into a single system. A survey~\cite{o20102010} shows that over 50\% of users at organizations with more than 500 employees are given some of their permissions to access resources based on RBAC. To adopt RBAC into a system, organizations often use hybrid approaches, which combine RBAC and access control lists since specific user types, systems, and workflow may not be effective to manage access based on roles. Therefore, when RBAC is extended and combined with other access control models, software developers often require to provide a formal representation, which can support of proof of the model. Then, software developers implement extended models based on NIST standard of RBAC implementation. Especially, software developers should understand intended use of extended access control models to meet a new system requirement. For example, for a spatial RBAC model, software developers incorporate additional spatial constraints, which can be either static or dynamic, in practice into the RBAC model.

>>>>

This standard does not provide a specific guideline for implementation of various extended RBAC models.
Moreover, when software developers implemented extended RBAC models, they require clear security requirements of extended role-based access control, which identifies security objectives, intended environments, and assumed threats. Moreover, software developers understand how this extended RBAC overlaps with other access control models when more than two access control models are integrated into a single system. A survey~\cite{o20102010} shows that over 50\% of users at organizations with more than 500 employees are given some of their permissions to access resources based on RBAC. To adopt RBAC into a system, organizations often use hybrid approaches, which combine RBAC and access control lists because specific user types, systems, and workflow may not be effective to manage access based on roles. Therefore, when RBAC is extended and combined with other access control models, software developers often require to provide a formal representation, which can support of proof of the model. Then, software developers implement extended models based on NIST standard of RBAC implementation. Especially, software developers should understand intended use of extended access control models to meet a new system requirement. For example, for a spatial RBAC model, software developers incorporate additional spatial constraints, which can be either static or dynamic, in practice into the RBAC model.

### 2013-05-08 paper/sections/intro.tex
Updates to address grammatical issues.

\textit{The goal of our work is to aid practitioners and researchers in choosing an extension to the RBAC$_{m}$, and in understanding
how extensions to RBAC$_{m}$ are evaluated by providing an assessment of RBAC extensions. We accomplished this through: 
establishment of a set of RBAC extension categories, an examination of the state of the art in evaluations of RBAC extension, and a breakdown of the 
motivations that have led to extensions to RBAC$_{m}$.} To accomplish this goal, we seek to answer following research questions:

>>>>

\textit{The goal of our work is to aid practitioners and researchers in choosing an extension to the RBAC$_{m}$, and in understanding
how extensions to RBAC$_{m}$ are evaluated by providing an assessment of RBAC extensions. We accomplished this through 
establishment of a set of RBAC extension categories, an examination of the state of the art in evaluations of RBAC extension, and a breakdown of the 
motivations that have led to extensions to RBAC$_{m}$.} To accomplish this goal, we seek to answer following research questions:

### 2013-05-08 paper/sections/process.tex
Updates to address grammatical issues.

We adopted and applied a systematic literature review following recommendations from Kitchenham's suggested processes \cite{kitchenham2007guidelines}.  The systematic literature review process was broken down into four stages and the rest of this section is broken down by each stage.  The stages were:

@@

For the first phase of our systematic literature review, we developed a search strategy for finding papers.  The search strategy was executed by an automated comprehensive search taking as input a set of academic search engines and a list of search terms.  The search was performed by applying each search term to each engine incrementally until the stopping criteria were met. Table \ref{tab:search_results} lists the four search engines along the left most column with the three search terms across the top row along with the papers for each criterion and engine combination.  The search algorithm was performed as follows:

@@

Question 1 was a definitive exclusion criterion as any paper that failed in the affirmative was rejected.  Questions 3 and 4 were marked as conditional includes given that they were connected in making a decision.  A paper that met question 3 but not 4, or met 4 but not 3 would be included since in some cases the real world examples served as research evaluations and without this conditional include the paper list size would be too small to be significant.

>>>>

We adopted and applied a systematic literature review following recommendations from Kitchenham's suggested processes \cite{kitchenham2007guidelines}.  The systematic literature review process was broken down into four stages and the rest of this section is broken down by each stage.  The stages were

@@

For the first phase of our systematic literature review, we developed a search strategy for finding papers.  The search strategy was executed by an automated comprehensive search taking as input a set of academic search engines and a list of search terms.  The search was performed by applying each search term to each engine incrementally until the stopping criteria were met. Table \ref{tab:search_results} lists the four search engines along the left most column with the three search terms across the top row along with the papers for each criterion and engine combination.  The search algorithm was performed as follows

@@

Question 1 was a definitive exclusion criterion as any paper that failed in the affirmative was rejected.  Questions 3 and 4 were marked as conditional includes given that they were connected in making a decision.  A paper that met question 3 but not 4, or met 4 but not 3 would be included because for some cases the real world examples served as research evaluations and without this conditional include the paper list size would be too small to be significant.
